Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-04-29 paper/manuscript.tex
Start merging a new dummy atom section  On branch sboresch
 Changes to be committed:
	modified:   manuscript.tex
	modified:   sb.bib

>>>>

\paragraph{Dummy atoms}

In practically all relative free energy simulations, regardless of the approach, SCST, SCDT, or DCDT, dummy atoms (non-interacting placeholder atoms) are required. We stress this point, as sometimes dummy atoms are viewed as something specific to SCST. In SC approaches, dummy atoms may even be required if the number of atoms between initial and final state of the system is identical, consider, e.g., tautomeric forms of a molecule.

The need for and role of dummy atoms is clearly visible in Fig.~\ref{fig:fig_topology}: One or more dummy atoms are attached to the physical molecule. A crucial requirement is that these atoms, typically connected through bonded energy terms (bond stretching, angle bending, dihedral angles) do not affect equilibrium and dynamical properties of the physical system. However, since a physical molecule + dummy atoms attached to it can or must be viewed as a modified physical molecule, the partition function of the molecules with and without dummy atoms will NOT be identical. In other words, a single free energy difference between two molecules A and B will be different from that of A(+D) and B(+D), where the (+D) indicates the presence of one or more dummy atoms needed so that the alchemical transformation can be carried out. When a dual coordinate approach is used, the noninteracting part of the respective other endstate is tied to the physical one, and will contribute to the partition function. By construction, relative free energy simulatins involve two analogous alchemical transformations, e.g., A(+D) -> B(+D) in gas phase and aqueous solution (RSFEs), or in the bound and unbound state (RBFEs). If dummy atoms are interacting with the physical system in such a way that $\mathsf{Z = Z(R) \times Z(R\text{-}D) \times Z(D)}$, where Z is the total partition function, Z(R) the partition function of the physical system, Z(D) the partition function contribution resulting from interactions between only the dummy atoms, and Z(R-D) the contribution resulting from the bonded connections between the physical atoms and dummy atoms, then their contributions (Z(D) and Z(R-D)) factor from the total partition function. As dummy atoms do not interact with the environment, their contribution, therefore, is identical in the two alchemical transformations required in the two legs of the relative transformation and will cancel exactly from the double free energy difference of interest.

While some early work on the correct treatment of dummy atoms, i.e., how to treat them so that their contribution to the partition function factors, has been available for years \cite{Boresch_2002,Shobana_2000,Wang_2012}, most workers seem to mostly turn off nonbonded interactions for dummy atoms, but leaving all bonded interactions intact \sbnote{To all: While I am quite sure that this is what most of us have been doing most of the time, I wonder whether this has been epxlicitly described anywhere?}. Since current force fields use multiple angle and torsion terms for each atom, there are often a large number of bonded energy terms connecting dummy atoms to the physical molecule. However, as already described in the early studies and repeated in a more recent study focusing on the correct treatment of dummy atoms \cite{Fleck_2021}, the desired factorization can only be guaranteed if any dummy atom connected to the physical molecule is held by exactly three (non-redundant) degrees of freedom. Naive application of the three-degree-of-freedom per dummy atom rule (for dummy atoms having at least one joint bonded term with the physical molecule), however, can result in simulations where the dummy part adopts unphysical configurations/conformations; Fleck et al. referred to this as “flapping”. Thus, depending on the details of the physical molecule -- dummy atom(s) junction, the three degree of freedoms need to be selected with care, or a carefully chosen additional redundant degree of freedom may be needed. In the latter case, the factorization of the partition function is not exact, but the overall error can still be made arbitrarily small. For the full details, we refer the reader to Ref.~\cite{Fleck_2021}.

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. However, DC approaches are also extremely prone to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest.

The influence from an incorrect treatment of dummy atoms (e.g., leaving all bonded terms intact) is difficult to estimate, though it may be small in many cases \cite{Fleck2021}. Nevertheless, some guidelines should be followed:
\begin{itemize}
  \item Obviously, a system with and without dummy atoms should have the same energetic minimum. Note that in larger groups of dummy atoms, redundant bonded terms involving only dummy atoms will be essential to avoid flapping, and the energy of the optimized geometry may be different from that of the physical system without dummy atoms. Thus, a \underline{necessary} condition is as follows: Optimize the geometry of the physical system (no dummy atoms); the final energy is the reference value. Optimize the geometry of the same molecule in the presence of dummy atoms; the final energy may well be different from the reference energy. Now delete the dummy atoms and recompute the energy without further minimization. The energy after the deletion of the dummy atoms must be identical to the reference energy. 
\item The above is only a \underline{necessary} criterion, and the dummy atoms may still influence the dynamical behavior of the physical molecule. Furthermore, overzealous removal of redundant energy terms may result in flapping of the dummy atoms. If one suspects this to be the case, then a short simulation in the gas phase should reveal flapping. Unfortunately, while straightforward in principle, both checks described are extremely hard to automate, which is essential in large RFEs campaigns. 
\item The correct treatment of dummy atoms can be validated by closing a suitable thermodynamic cycle.  E.g., Fleck et al. calculated ASFEs for each of their endstates, as well as the RSFE between them. In all cases when dummy atoms were treated according to best practices, the cycle closure error was well within the statistical uncertainty, which was well below 0.1 kcal/mol. A related, earlier study by Loeffler et al. also reported ASFEs and RSFEs. If one attempts to close the cycle based on the numbers reported there, cycle closure error was nonnegligible in some cases. [Loeffler2017]
\item While solvation free energies can nowadays be computed quite fast, their systematic calculation would nevertheless be too costly. The role of solvent can be replaced by some external restraint, distorting the average geometry of the physical molecule. If one computes the free energy cost of applying the restraint with and without dummy atoms, the two numbers must agree within statistical significance. While much faster than full solvation free energy calculations, applying a restraint in the gas phase requires careful thermalization of the simulation system, and automation is difficult.
  \end{itemize}
Further work automating the correct setup and treatment of dummy atoms is necessary.
